Estimating the dynamics of SPARQL query results using binary classification

### 2019-08-06 main.tex
Camera ready draft

Given a SPARQL query and a dynamic RDF graph (consisting of multiple historical versions), in this work, we address the problem of predicting whether or not the query's results will change in the next version of the RDF graph. To the best of our knowledge, this is the first work to address this problem. Along these lines, we evaluate a general architecture based on Machine Learning that extracts static features from a query, as well as dynamic features from the query and dataset. These features are fed into a binary classifier to predict whether or not the query results will change in a fixed point in the future, or, more ambitiously, can be fed into a regression model to predict when the query results are likely to change. With respect to the features used, we show that there is a trade-off between those that are easy to compute but offer less accurate predictions (e.g., static query features) versus those that are more costly to compute but offer more accurate predictions (e.g., historical changes in query results). Per this trade-off, which features to use for predicting the dynamics of query results may then depend on the particular application.

@@

A variety of works have addressed the issue of modeling and consuming dynamic Linked Data from a broad range of perspectives~\cite{PassantM10,TrampFEA10,UmbrichHHPD10,UmbrichKL10,KaferAUOH13,DividinoSGG13,DividinoKG14,Kjernsmo15,NishiokaS16,GonzalezH18}. One of the major challenges considered is that of keeping cached copies of remote dynamic data -- cached for reasons of efficiency and scalability -- up-to-date on the consumer side, which we refer to as the \textit{synchronization problem}.

Some works have addressed the synchronization problem on the publisher side, proposing notification mechanisms that keep registered consumers informed about relevant changes to the data~\cite{MissierACDG07,PassantM10,TrampFEA10,MaderMS14,webSub18}; although such approaches may facilitate strong consistency -- meaning that consumers are kept up-to-date with the remote data on the publisher side -- they centralize the burden  of synchronization on the publisher, potentially leading to scalability issues.

@@

SPARQL is the recommended query language to retrieve and manipulate data stored in the RDF format. In this work, we focus on SPARQL \texttt{SELECT} queries, where we will first define a SPARQL 1.0 query. Let $\Vars$ be a set of variables disjoint from the set of RDF terms. A \emph{SPARQL expression} is built recursively as follows. (1) A \emph{triple pattern} $t \in  (\Iris \cup \Bnodes \cup \Lits \cup \Vars) \times (\Iris \cup \Vars) \times (\Iris \cup \Lits \cup \Bnodes \cup \Vars)$ is an \emph{expression}. (2) If $Q_1$ and $Q_2$ are expressions and $R$ is a filter condition, then $Q_1$ \texttt{FILTER} $R$, $Q_1$ \texttt{UNION} $Q_2$, $Q_1$ \texttt{OPTIONAL} $Q_2$, $Q_1$ \texttt{AND} $Q_2$ are $expressions$. Finally, if $Q$ is an expression, $V$ a list of variables and $\Delta$ a boolean value, $\texttt{SELECT}_V^\Delta Q$ is a SPARQL \texttt{SELECT} query, where $V$ denotes the projected variables, and $\Delta$ the \texttt{DISTINCT} option that when true, removes duplicate results. The semantics of a SPARQL \texttt{SELECT} query $Q$ is defined in terms of its \emph{evaluation} over an RDF graph $G$, denoted $Q(G)$, giving a set of partial mappings from projected variables to the set $\Iris \cup \Lits \cup \Bnodes$; we refer to P{\'{e}}rez et al.~\cite{PerezAG09} for definitions. (According to the standard, a SPARQL query is evaluated over a set of \emph{named graphs}; however, since the Wikidata query service indexes data in one RDF graph, to keep discussion concise, we consider RDF graphs, though our methods can be generalized in a straightforward manner to the named graphs setting.)

@@

\centering
\resizebox{0.9\textwidth}{!}{
\begin{tikzpicture}
\node[i] (q) {$Q$};

@@

\node[c,right=1.5\hgap of f6.south,minimum width=6.5em] (c) {Classification\hsp};

@@

The first group of features captures statistics about the query, indicating loosely its size and complexity. One might consider that the higher these values are, in general, the more dynamic we can expect the query to be since there are more ``opportunities'' for the query results to be affected by change. In some cases, however, the hypothesized correlation is not direct since, for example, adding more triple patterns may serve to narrow the query down and focus it on a static part of the graph (e.g., when looking for the movies of directors, adding a triple pattern to restrict the results to directors who have died may reduce the likelihood of changes in the results in a future version). Hence it will be of interest to see how these features affect the predictions made experimentally.

@@

In the third group gathers related features into one dimension: in the case of \textit{Recursive path}, we group queries with path expressions of the form $e$\texttt{*} or $e$\texttt{+}. On the other hand, in \textit{Negation}, we group non-monotonic features that allow for modeling difference (\texttt{MINUS}, \texttt{NOT EXISTS}, \texttt{!BOUND}). These features -- though course-grained -- are straightforward to extract from a query, and offer valuable insights into how the query may behave in a dynamic query; for example, the \textit{Negation} feature captures information about the (non-)monotonicity of the query, while we suppose that \textit{Recursive paths}, which may traverse an arbitrary number of triples in the graph, might be more sensitive to change. Again, such correlations are not without exception and will require empirical study.

@@

\paragraph{Predicate Features:} The next component extracts features that capture information about the dynamics of predicates used in the query (without \textit{evaluating} the query on the dynamic graph). This component captures how often and how many triples change for a predicate in a time interval, with the idea that -- as in previous works~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12,DehghanzadehPKUHD14} -- predicates capture rich information about the dynamics of a dataset, where the results of queries with dynamic predicates will be more sensitive to change. Furthermore, given that the number of predicates is often \textit{relatively} low (in the thousands or tens of thousand) for most datasets, such statistics are quite lightweight to maintain and compute. Formally, given two RDF graphs $G_i$ and $G_j$, we denote by $G_i \oplus G_j$ the set of triples $(G_i \cup G_j) - (G_i \cap G_j)$ where ``$-$'' denotes set difference; in other words, $G_i \oplus G_j$ denotes the triples in one graph or the other but not both (noting that $G_i \oplus G_j = G_j \oplus G_i$). Next, given an RDF graph $G$ and an IRI $p$, let $\#(G,p) \da |\{(x,y,z) \in G : y = p \}|$ denote the number of triples in $G$ with predicate $p$. Finally, given a dynamic RDF graph $\mathcal{G} \da ( G_1, \ldots, G_n )$ and a predicate $p$, we denote by $\Delta(\mathcal{G},p) \da \Sigma_{i=1}^{n-1}\frac{\#(G_i \oplus G_{i+1},p)}{\#(G_i \cup G_{i+1},p)}$ the normalized sum of the number of triples with the predicate $p$ that changed between pairs of consecutive versions.

@@

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-score for the predictions made considering the union of different combinations of the sets of features (\textsc{q}, \textsc{p}, \textsc{r}) and varying the window sizes (3, 5, 9, 17); for reference, we include a baseline that randomly guesses \texttt{yes}/\texttt{no}. Given the number of configurations presented, we only include $F_1$ scores for reasons of space; however, we remark that the Precision and Recall scores were in general quite balanced. The best results were obtained using the features from \textsc{r}, where, as can be expected, prediction performs best when knowledge of changes in the historical results of the input query is available. The features from sets \textsc{q} and \textsc{p} -- which do not assume the availability of such information -- obtain a significantly lower $F_1$-score; the best result including \textsc{r} was with \textsc{pr} ($F_1 = 0.831$) using Linear SVM and window size 17, while the best results without \textsc{r} was with \textsc{q} ($F_1 = 0.60$) using Linear SVM and window size 5.

@@

In this paper we evaluate methods to predict whether or not (OSC) the results of an input query will change in a fixed point in the near future. More specifically, we evaluate a framework based on binary classifiers that accept features extracted from the query, from past versions of the data, and/or from the combination of both. Considering this framework, we hypothesize that there is a conceptual trade-off between the cost of computing features and their value for prediction: features extracted from queries alone are the most efficient to extract, not requiring historical data, but are quite coarse-grained for prediction; on the other hand, features based on changes in results over previous versions for an (unseen) input query are often the most costly to acquire, but offer fine-grained information for prediction; finally, features based on the dynamics of predicates offer a balance between the two, allowing to summarize historical data into succinct statistics, thus offering more fine-grained information than static query features but more coarse-grained information than historical results.

We thus explore this trade-off experimentally. For this, we define some desirable characteristics for the queries and data used in the experimental setting, such as availability, size, dynamics, diversity, etc. Settling on Wikidata, we collect 23 versions and curate 221 user-generated queries from which we compute results, statistics and ultimately features that form the gold standard for the tasks. Analysing these data and queries, we conclude that they meet the desired characteristics, offering a large-scale, diverse, real-world setting; more importantly, we find that our queries offer an ideal distribution of queries whose results always change, queries whose results never change, and (more importantly) queries whose results change intermittently and for which prediction is thus non-trivial. We make this gold standard available at \url{https://users.dcc.uchile.cl/~amoya/quweda2019/}.

Finally, we use this gold standard dataset to evaluate the trade-off identified between different types of features for OSC predictions. For OSC, we evaluate the success rates of a variety of classifiers based on the $F_1$-measure of the predictions. Our results show that the features based on historical changes to query results perform best (reaching $F_1 = 0.831$ in the best configuration), whereas considering static query features and predicate dynamics alone is less competitive (reaching $F_1 =0.600$ in the best configuration) when historical results are not available or are prohibitively costly to compute. Comparing query and predicate features, in fact, query features performed better, where predicate features alone only barely outperformed a random baseline.

As a conclusion, our results show that having knowledge of the historical changes of the results of a query is important for improving the quality of OSC predictions using binary classifiers. However, in many scenarios, such knowledge is often not available or may not be practical to compute. Considering a real-world caching use-case, for example, historical changes in results may be maintained for queries that are frequently repeated with relatively low-cost, but for a previously unseen query, computing results for past versions at runtime would incur a prohibitive cost. Hence, we identify an open research question: is it possible to estimate the dynamics of query results without relying on (costly) knowledge about historical changes of query results while staying competitive with the quality of prediction possible when such knowledge is available?

In terms of other future work, gold standards based on datasets other than Wikidata could be pursued. Also, our gold standard based on Wikidata could be extended to consider more versions spanning a longer period of time and/or more queries; a very large query dataset for Wikidata was recently published~\cite{MalyshevKGGB18} from which it should be possible to extend the Wikidata gold standard (the challenge being to keep a balance of static and dynamic queries with diverse features). Concerning the prediction tasks themselves, there are a variety of other features and approaches that could be applied; one promising direction may be, for example, to apply a more fine-grained analysis of the query, considering for example estimates of the selectivity of particular triples patterns, and combining this information with statistics on the dynamics of the data and the query operators used; our current gold standard could be used to validate such developments. Admittedly, there may also be better ways to perform such predictions without relying on binary classifiers, but rather using more analytical methods. Finally, it would be interesting to investigate the effectiveness of these techniques in practice, developing caching systems, synchronization schedules, and other applications, based on the evaluated predictions.

%\paragraph{Acknowledgements} This work was supported by the Millennium Institute for Foundational Research on Data (IMFD) and by Fondecyt Grant No. 1181896.

>>>>

Given a SPARQL query and a dynamic RDF graph (consisting of multiple historical versions), in this work, we address the problem of predicting whether or not the query's results will change in the next version of the RDF graph. To the best of our knowledge, this is the first work to address this problem. Along these lines, we evaluate a general architecture based on Machine Learning that extracts static features from a query, as well as features from the query and dynamic dataset. These features are fed into a binary classifier to predict whether or not the query results will change in a fixed point in the future, or, more ambitiously, can be fed into a regression model to predict when the query results are likely to change. With respect to the features used, we show that there is a trade-off between those that are easy to compute but offer less accurate predictions (e.g., static query features) versus those that are more costly to compute but offer more accurate predictions (e.g., historical changes in query results). Per this trade-off, which features to use for predicting the dynamics of query results may then depend on the particular application.

@@

A variety of works have addressed the issue of modeling and consuming dynamic Linked Data from a broad range of perspectives~\cite{PassantM10,TrampFEA10,UmbrichHHPD10,UmbrichKL10,KaferAUOH13,DividinoSGG13,IbanezSMC14,DividinoKG14,Kjernsmo15,NishiokaS16,GonzalezH18}. One of the major challenges considered is that of keeping cached copies of remote dynamic data -- cached for reasons of efficiency and scalability -- up-to-date on the consumer side, which we refer to as the \textit{synchronization problem}.

Some works have addressed the synchronization problem on the publisher side, proposing notification mechanisms that keep registered consumers informed about relevant changes to the data~\cite{MissierACDG07,PassantM10,TrampFEA10,MaderMS14,IbanezSMC14,webSub18}; although such approaches may facilitate strong consistency -- meaning that consumers are kept up-to-date with the remote data on the publisher side -- they centralize the burden  of synchronization on the publisher, potentially leading to scalability issues.

@@

SPARQL is the recommended query language to retrieve and manipulate data stored in the RDF format. In this work, we focus on SPARQL \texttt{SELECT} queries, where we will first define a SPARQL 1.0 query. Let $\Vars$ be a set of variables disjoint from the set of RDF terms. A \emph{SPARQL expression} is built recursively as follows. (1) A \emph{triple pattern} $t \in  (\Iris \cup \Bnodes \cup \Lits \cup \Vars) \times (\Iris \cup \Vars) \times (\Iris \cup \Lits \cup \Bnodes \cup \Vars)$ is an \emph{expression}. (2) If $Q_1$ and $Q_2$ are expressions and $R$ is a filter condition, then $Q_1$ \texttt{FILTER} $R$, $Q_1$ \texttt{UNION} $Q_2$, $Q_1$ \texttt{OPTIONAL} $Q_2$, $Q_1$ \texttt{AND} $Q_2$ are $expressions$. Finally, if $Q$ is an expression, $V$ a list of variables and $\Delta$ a boolean value, $\texttt{SELECT}_V^\Delta Q$ is a SPARQL \texttt{SELECT} query, where $V$ denotes the projected variables, and $\Delta$ the \texttt{DISTINCT} option that when true, removes duplicate results. The semantics of a SPARQL \texttt{SELECT} query $Q$ is defined in terms of its \emph{evaluation} over an RDF graph $G$, denoted $Q(G)$, giving a set of partial mappings from projected variables to the set $\Iris \cup \Lits \cup \Bnodes$; we refer to P{\'{e}}rez et al.~\cite{PerezAG09} for definitions. (A SPARQL query is in fact evaluated over a set of \emph{named graphs}; though we consider RDF graphs here, our methods also generalize to the named graphs setting.)

@@

\centering
\resizebox{0.8\textwidth}{!}{
\begin{tikzpicture}
\node[i] (q) {$Q$};

@@

\node[c,right=2\hgap of f6.south,minimum width=6.5em] (c) {Classification\hsp};

@@

The first group of features captures statistics about the query, indicating loosely its size and complexity. One might consider that the higher these values are, in general, the more dynamic we can expect the query to be since there are more ``opportunities'' for the query results to be affected by change. In some cases, however, the hypothesized correlation is not direct since, for example, adding more triple patterns may serve to narrow the query down and focus it on a static part of the graph (e.g., when looking for the movies of directors, adding a triple pattern to restrict the results to directors who have died may reduce the likelihood of changes in the results in a future version). Hence it will be of interest to see, experimentally, how these features affect the predictions.

@@

In the third group gathers related features into one dimension: in the case of \textit{Recursive path}, we group queries with path expressions of the form $e$\texttt{*} or $e$\texttt{+}. On the other hand, in \textit{Negation}, we group non-monotonic features that allow for modeling difference (\texttt{MINUS}, \texttt{NOT EXISTS}, \texttt{!BOUND}\footnote{We recall that \texttt{!BOUND} can be combined with \texttt{OPTIONAL} to express negation.}. These features -- though course-grained -- are straightforward to extract from a query, and offer valuable insights into how the query may behave in a dynamic query; for example, the \textit{Negation} feature captures information about the (non-)monotonicity of the query, while we suppose that \textit{Recursive paths}, which may traverse an arbitrary number of triples in the graph, might be more sensitive to change. Again, such correlations are not without exception and will require empirical study.

@@

\paragraph{Predicate Features:} The next component extracts features that capture information about the dynamics of predicates used in the query (without \textit{evaluating} the query on the dynamic graph). This component captures how often and how many triples change for a predicate in a time interval, with the idea that -- as in previous works~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12,DehghanzadehPKUHD14} -- predicates capture rich information about the dynamics of a dataset, where the results of queries with dynamic predicates will be more sensitive to change. Assuming that the number of predicates is \textit{relatively} low, such statistics can be easier to maintain. Formally, given two RDF graphs $G_i$ and $G_j$, we denote by $G_i \oplus G_j$ the set of triples $(G_i \cup G_j) - (G_i \cap G_j)$ where ``$-$'' denotes set difference; in other words, $G_i \oplus G_j$ denotes the triples in one graph or the other but not both (noting that $G_i \oplus G_j = G_j \oplus G_i$). Next, given an RDF graph $G$ and an IRI $p$, let $\#(G,p) \da |\{(x,y,z) \in G : y = p \}|$ denote the number of triples in $G$ with predicate $p$. Finally, given a dynamic RDF graph $\mathcal{G} \da ( G_1, \ldots, G_n )$ and a predicate $p$, we denote by $\Delta(\mathcal{G},p) \da \Sigma_{i=1}^{n-1}\frac{\#(G_i \oplus G_{i+1},p)}{\#(G_i \cup G_{i+1},p)}$ the normalized sum of the number of triples with the predicate $p$ that changed between pairs of consecutive versions.

@@

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-score for the predictions made considering the union of different combinations of the sets of features (\textsc{q}, \textsc{p}, \textsc{r}) and varying the window sizes (3, 5, 9, 17); for reference, we include a baseline that randomly guesses \texttt{yes}/\texttt{no} respecting the observed class distribution. Given the number of configurations presented, we only include $F_1$ scores for reasons of space; however, we remark that the Precision and Recall scores were in general quite balanced. The best results were obtained using the features from \textsc{r}, where, as can be expected, prediction performs best when knowledge of changes in the historical results of the input query is available. The features from sets \textsc{q} and \textsc{p} -- which do not assume the availability of such information -- obtain a significantly lower $F_1$-score; the best result including \textsc{r} was with \textsc{pr} ($F_1 = 0.831$) using Linear SVM and window size 17, while the best results without \textsc{r} was with \textsc{q} ($F_1 = 0.60$) using Linear SVM and window size 5.

@@

In this paper we evaluate methods to predict whether or not (OSC) the results of an input query will change at a fixed point in the near future. More specifically, we evaluate a framework based on binary classifiers that accept features extracted from the query, from past versions of the data, and/or from the combination of both. Considering this framework, we hypothesize that there is a conceptual trade-off between the cost of computing features and their value for prediction: features extracted from queries alone are the most efficient to extract, not requiring historical data, but are quite coarse-grained for prediction; on the other hand, features based on changes in results over previous versions for an (unseen) input query are often the most costly to acquire, but offer fine-grained information for prediction; finally, features based on the dynamics of predicates offer a balance between the two, allowing to summarize historical data into succinct statistics, thus offering more fine-grained information than static query features but more coarse-grained information than historical results.

We thus explore this trade-off experimentally using 23 versions and 221 user-contributed queries for Wikidata to form a gold standard dataset. We use this gold standard dataset to evaluate the trade-off identified between different types of features for OSC predictions. Our results show that the features based on historical changes to query results perform best (reaching $F_1 = 0.831$ in the best configuration), whereas considering static query features and predicate dynamics alone is less competitive (reaching $F_1 =0.600$ in the best configuration) when historical results are not available or are prohibitively costly to compute. Comparing query and predicate features, in fact, query features performed better, where predicate features alone only barely outperformed a random baseline.

In conclusion, our results show that having knowledge of the historical changes of the results of a query is important for improving the quality of OSC predictions using binary classifiers. However, in many scenarios, such knowledge is often not available or may not be practical to compute. Considering a real-world caching use-case, for example, historical changes in results may be maintained for queries that are frequently repeated with relatively low-cost, but for a previously unseen query, computing results for past versions at runtime would incur a prohibitive cost. Hence, we identify an open research question: is it possible to estimate the dynamics of query results without relying on (costly) knowledge about historical changes of query results while staying competitive with the quality of prediction possible when such knowledge is available?

Regarding future work, our gold standard based on Wikidata could be extended to consider more versions spanning a longer period of time and/or more queries; a very large query dataset for Wikidata was recently published~\cite{MalyshevKGGB18}, which may be of use here. Building gold standards based on other datasets would also help to diversify the evaluation process. Concerning the prediction tasks themselves, one promising direction may be to apply a more fine-grained analysis of the query, considering (for example) the selectivity of particular triples patterns as well as their dynamicity. There may also be better ways to perform such predictions without relying on binary classifiers, but rather using more analytical methods. Finally, it would be interesting to investigate the effectiveness of these techniques in practice, developing caching systems, synchronization schedules, and other applications, based on the evaluated predictions.

\paragraph{Material} We make supplementary material (queries, data, results, etc.) available at \url{https://users.dcc.uchile.cl/~amoya/quweda2019/}.

\paragraph{Acknowledgements} This work was supported by the Millennium Institute for Foundational Research on Data (IMFD) and by Fondecyt Grant No. 1181896.
